```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## load radiant packages if neededi
if (!exists("r_environment")) library(radiant)
```

```{r}
install.packages('countrycode')
library(countrycode)
```

```{r}  
## read in data
data<-read.csv("suicide_rates.csv")
```

```{r}
data$continent <- countrycode(sourcevar = data$country,
                                  origin = 'country.name',
                                  destination = 'continent')

data$age <- factor(data$age, levels = c('5-14 years', '15-24 years', '25-34 years', 
                                                   '35-54 years', '55-74 years', '75+ years'))
data$sex <- factor(data$sex, levels = c('male', 'female'))
data$continent <- countrycode(sourcevar = data$country,
                                  origin = 'country.name',
                                  destination = 'continent')
data <- data %>%
  filter(year < 2016)
str(data)
```

```{r fig.width = 9.15, fig.height = 9.15, dpi = 144}
result <- correlation(
  data, 
  vars = c(
    "sex", "suicides_no", "population", "suicides.100k.pop", 
    "gdp_for_year....", "gdp_per_capita....", "continent"
  )
)
summary(result)
plot(result, nrobs = 1000)
```

In these resultes, we could see that most of them are significant.

The Pearson correlation between `suicides_no` and `population` is 0.62, which indicates that there is a moderate positive relationship between them. It makes sense because the number of suicides will to some extent increase as the total number of popultation increse. 

However, when we look at the relationsihp between `suicides.100k.pop` and `population`, it shows no linear relationship. The conclusion is that the average suicide rate would not necessarily increase as the population become larger.









